\documentclass[output=paper]{langscibook}

\author{Claire Bowern\affiliation{Yale University}}
\title{Theoretical issues in computational approaches to semantic change} 
\abstract{
Computational approaches to semantic change are predicated on a set of theoretical assumptions about the nature of meaning, the distribution of lexical items, and what it means to identify differences between two distributions over time. In this chapter, we will provide commentary on some of these key assumptions and their relationships to both data and to methods of modeling language change. We will show that word embeddings, while showing types of short-term change, have numerous problems that make them unreliable for fine-grained claims.
 
Secondly, the chapter will sketch an approach to the study of semantic change which follows from evolutionary principles. Models of cultural and linguistic evolution that work on Darwinian principles include mechanisms for transmission and selection of traits which exhibit variation. Following other work in linguistics, we will show that these principles are satisfied by linguistic systems. Viewing change from this viewpoint allows us to constrain a typology of semantic change and make predictions about the mechanisms which are likely to be active in how the lexicon evolves over time. In contrast, methods which simply compare slices of the lexicon at different points in time, uncontrolled for corpus composition changes, genre, or other types of skewing, are unlikely to give us a clear picture in phylogenetic time.
}

\begin{document}
\maketitle

%\section{Introduction}
 
 
%\section*{Abbreviations}
%\begin{tabularx}{.45\textwidth}{lQ}
%... & \\
%... & \\
%\end{tabularx}
%\begin{tabularx}{.45\textwidth}{lQ}
%... & \\
%... & \\
%\end{tabularx}


%\section*{Acknowledgements}
%\citet{Nordhoff2018} is useful for compiling bibliographies.

{\sloppy\printbibliography[heading=subbibliography,notkeyword=this]}
\end{document}
